Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-05-01 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Add brief captions

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLL}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \citet{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \citet{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \citet{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabulary}
\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
\label{tab:bs-seq_mappers}
\end{table}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
\caption{M-bias plot for the \emph{ADS} sample from the \emph{Lister} dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.}
\label{fig:ADS_M-bias}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately. For CpGs in \emph{read-1}, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in \emph{read-2}, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read-positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in \emph{read-1}. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\label{fig:E18VA_M-bias}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveals the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Chromosome & Strand & Position & M & U \\ \midrule
chr1       & $+$      & 100      & 7 & 1 \\
chr1       & $-$      & 101      & 5 & 2 \\
chr2       & $+$      & 400      & 0 & 3 \\
chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
\end{tabular}
\caption{Example of output for methylation calling at 1-tuples. Each row records the the number of methylated ($M$) and unmethylated ($U$) reads at a 1-tuple. Loci may be stratified by strand, as is done here, in which case most CpGs will have measurements for both the positive and negative strands.}
\label{tab:1-tuples}
\end{table}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
\caption{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
\label{fig:incomplete_picture_1-tuples}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLL}
\toprule
Software           & Reference                                & Input                                              & Limitations                                                                \\ \midrule
methclone & \citet{Li:2014ei}                       & \texttt{Bismark} \texttt{BAM}                                  & Unable to install \\
methpat  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bme} & Designed for amplicons not whole-genome data.                                                                                 \\
DMEAS    & \citet{He:2013cj}                       & Output of \texttt{bme} & Windows operating system only. Perl code only available as PDF file.             \\
\bottomrule
\end{tabulary}
\caption{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}). The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
\label{fig:methtuple_example1}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_memory}
\end{figure}

>>>>

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLL}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \citet{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \citet{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \citet{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabulary}
\caption[Popular bisulfite-sequencing read mappers and their underlying mapping software]{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
\label{tab:bs-seq_mappers}
\end{table}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
\caption[\emph{ADS} M-bias]{M-bias plot for the \emph{ADS} sample from the \emph{Lister} dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately.}
\label{fig:ADS_M-bias}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption[\emph{E18VA} M-bias]{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately. For CpGs in \emph{read-1}, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in \emph{read-2}, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read-positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in \emph{read-1}. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\label{fig:E18VA_M-bias}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption[\texttt{Bis-SNP} method schematic]{\texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveals the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Chromosome & Strand & Position & M & U \\ \midrule
chr1       & $+$      & 100      & 7 & 1 \\
chr1       & $-$      & 101      & 5 & 2 \\
chr2       & $+$      & 400      & 0 & 3 \\
chr2       & $+$      & 450      & 1 & 2  \\ \bottomrule
\end{tabular}
\caption[Methylation patterns at 1-tuples]{Example of output for methylation calling at 1-tuples. Each row records the the number of methylated ($M$) and unmethylated ($U$) reads at a 1-tuple. Loci may be stratified by strand, as is done here, in which case most CpGs will have measurements for both the positive and negative strands.}
\label{tab:1-tuples}
\end{table}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
\caption[Different methylation patterns can lead to identical $\beta$-values]{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
\label{fig:incomplete_picture_1-tuples}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLL}
\toprule
Software           & Reference                                & Input                                              & Limitations                                                                \\ \midrule
methclone & \citet{Li:2014ei}                       & \texttt{Bismark} \texttt{BAM}                                  & Unable to install \\
methpat  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bme} & Designed for amplicons not whole-genome data.                                                                                 \\
DMEAS    & \citet{He:2013cj}                       & Output of \texttt{bme} & Windows operating system only. Perl code only available as PDF file.             \\
\bottomrule
\end{tabulary}
\caption[Software for calling methylation patterns at m-tuples]{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bme} =  \texttt{bismark\_methylation\_extractor}.}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption[\texttt{methtuple} example region]{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}). The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
\label{fig:methtuple_example1}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption[\texttt{methtuple} 1-tuples]{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption[\texttt{methtuple} 3-tuples]{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption[\texttt{methtuple} 3-tuples with \texttt{--all-combinations}]{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption[Number of CpGs per read in \emph{Lister} dataset]{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption[\texttt{methtuple} 3-tuples example output]{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption[\texttt{methtuple} memory usage]{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_memory}
\end{figure}
